---
title: "Data Preparation"
toc: true
toc-location: right
toc-depth: 2
page-layout: full
number-sections: true
format:
  html:
    code-fold: true
    code-tools: false
execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

<style>
/* Keep the page table of contents readable even when custom site CSS is used. */
#quarto-margin-sidebar {
  min-width: 17rem !important;
  width: 17rem !important;
}

#TOC,
#TOC ul,
#TOC li,
#TOC a {
  line-height: 1.45 !important;
}

#TOC li {
  height: auto !important;
  margin-bottom: 0.28rem !important;
}

#TOC a {
  display: block;
  white-space: normal !important;
}

/* Keep wide preparation tables inside the page instead of shrinking the text. */
.cell-output-display,
.cell-output {
  max-width: 100%;
  overflow-x: auto;
}

#quarto-document-content table {
  width: 100%;
  font-size: 0.82rem;
}

#quarto-document-content th,
#quarto-document-content td {
  vertical-align: top;
}
</style>

This page documents how the VAST Challenge 2026 Mini-Challenge 1 data, stored as a nested RDS object, is converted into clean, reusable tables for the investigation and the Shiny application. The RDS object was previously created from the supplied JSON data. This page focuses only on data loading, validation and feature engineering; no charts are produced here.

# Setup

## Loading packages

The following packages are used to load and standardise the nested RDS source, prepare tabular data, derive time-based and case-specific features, and create network-ready data tables.

```{r}
#| echo: false
pacman::p_load(
  jsonlite,
  tidyverse,
  lubridate,
  knitr,
  tidygraph,
  igraph,
  readr
)
```

```{r}
package_summary <- tribble(
  ~Package, ~Purpose,
  "jsonlite", "Standardises the nested RDS structure into table-ready data.",
  "tidyverse", "Supports data transformation, reshaping and text preparation.",
  "lubridate", "Converts timestamps and derives temporal features.",
  "knitr", "Formats reproducible data tables in Quarto.",
  "tidygraph / igraph", "Creates network-ready node and edge structures.",
  "readr", "Loads the source RDS file and saves the prepared data object for later reuse."
)

knitr::kable(package_summary)
```

```{r}
#| echo: false

# Produces compact preview tables for long text-based MC1 fields.
show_table <- function(data, n = 10, text_columns = character(), max_chars = 110) {
  preview <- data %>%
    slice_head(n = n) %>%
    mutate(
      across(
        any_of(text_columns),
        ~ stringr::str_trunc(as.character(.x), width = max_chars)
      )
    )

  knitr::kable(preview, format = "html")
}
```

## Loading the raw MC1 data

The supplied Mini-Challenge 1 dataset has already been converted from JSON into an RDS object named `mc1_data.rds`. It contains a sequence of simulation rounds, each with participants, communication records and environmental context. The RDS object is loaded using a relative path so that the project can be rendered on another group member's computer without changing the code.

```{r}
# This QMD file is stored in the Methodology folder.
# Therefore, the source data is referenced relative to this folder.
data_path <- file.path("data", "MC1", "mc1_data.rds")

if (!file.exists(data_path)) {
  stop(
    "Dataset not found. Please save mc1_data.rds in: ",
    normalizePath(file.path("data", "MC1"), winslash = "/", mustWork = FALSE)
  )
}

# Load the nested RDS object that was previously created from the MC1 JSON file.
mc1_rds <- readr::read_rds(data_path)

if (!is.list(mc1_rds) || is.null(mc1_rds$rounds)) {
  stop(
    "mc1_data.rds was loaded, but it does not contain the expected $rounds element. ",
    "Please ensure that this is the MC1 RDS file created from MC1_final_00.json."
  )
}

# The source RDS stores deeply nested lists. This standardisation step keeps the
# existing unnesting code below compatible with the RDS version of the data.
mc1_data <- jsonlite::fromJSON(
  jsonlite::toJSON(mc1_rds, auto_unbox = TRUE, null = "null"),
  simplifyVector = TRUE
)

df_rounds_raw <- tibble::as_tibble(mc1_data$rounds)

round_summary <- tibble(
  Measure = c("Number of simulation rounds", "Fields in each round"),
  Value = c(nrow(df_rounds_raw), ncol(df_rounds_raw))
)

knitr::kable(round_summary)
```

# Extracting the nested tables

The data is stored in nested structures. It is unpacked into separate tables so that participant activity, communications and contextual events can be analysed without duplicating records.

## Participants

Each participant record is retained with the simulation round in which it appeared. Any available agent metadata is unpacked into separate columns.

```{r}
df_participants <- df_rounds_raw %>%
  select(hour, participants) %>%
  unnest(participants) %>%
  unpack(any_of("agent_round_metadata"), names_sep = "_")

participants_preview <- df_participants %>%
  select(any_of(c("hour", "agent_id", "agent_role", "agent_label", "declared_action")))

show_table(participants_preview)
```

## Communications

Communication records are unpacked at message level. Internal state and agent-round metadata are retained because they may later support behavioural and governance analysis.

```{r}
df_communications <- df_rounds_raw %>%
  select(hour, communications) %>%
  unnest(communications) %>%
  unpack(any_of(c("internal_state", "agent_round_metadata")), names_sep = "_")

normalise_recipients <- function(x) {
  if (is.null(x) || length(x) == 0 || all(is.na(x))) {
    return(NA_character_)
  }

  stringr::str_c(as.character(unlist(x, use.names = FALSE)), collapse = ", ")
}

communications <- df_communications %>%
  mutate(
    recipients = purrr::map_chr(recipients, normalise_recipients)
  ) %>%
  rename(round_hour = hour)

communications_preview <- communications %>%
  select(any_of(c(
    "round_hour", "message_id", "agent_id", "agent_role", "channel",
    "recipients", "content", "timestamp"
  )))

show_table(
  communications_preview,
  text_columns = c("recipients", "content"),
  max_chars = 95
)
```

## Environmental context

The scalar fields in the environmental context, including the market snapshot, are retained in one table. Nested event arrays are extracted separately because a single simulation round can contain several events.

```{r}
df_environment <- df_rounds_raw %>%
  select(hour, environment_context) %>%
  unpack(environment_context, names_sep = "_") %>%
  unpack(any_of("environment_context_market_snapshot"), names_sep = "_") %>%
  select(where(~ !is.list(.)))

# Derive clean fields where the corresponding source columns are available.
df_environment_clean <- df_environment %>%
  mutate(hour_datetime = lubridate::ymd_hms(hour, quiet = TRUE))

if ("environment_context_market_snapshot_stock_price" %in% names(df_environment_clean)) {
  df_environment_clean <- df_environment_clean %>%
    mutate(
      stock_price_numeric = readr::parse_number(
        as.character(environment_context_market_snapshot_stock_price)
      )
    )
}

if ("environment_context_market_snapshot_percent_change" %in% names(df_environment_clean)) {
  df_environment_clean <- df_environment_clean %>%
    mutate(
      stock_percent_numeric = readr::parse_number(
        as.character(environment_context_market_snapshot_percent_change)
      )
    )
}

environment_preview <- df_environment_clean %>%
  select(any_of(c(
    "hour", "environment_context_event_narrative",
    "environment_context_event_headline",
    "environment_context_market_snapshot_stock_price",
    "environment_context_market_snapshot_percent_change",
    "stock_price_numeric", "stock_percent_numeric"
  )))

show_table(
  environment_preview,
  text_columns = c(
    "environment_context_event_narrative",
    "environment_context_event_headline"
  ),
  max_chars = 120
)
```

## Environmental event tables

The helper function below extracts each nested environment array and keeps the `hour` field as the key for joining events back to a simulation round.

```{r}
#| echo: false
extract_environment_array <- function(rounds, array_column) {

  environment_data <- rounds %>%
    select(hour, environment_context) %>%
    unpack(environment_context, names_sep = "_")

  if (!array_column %in% names(environment_data)) {
    return(tibble(hour = character()))
  }

  environment_data %>%
    select(hour, all_of(array_column)) %>%
    unnest(all_of(array_column), keep_empty = FALSE)
}

df_media_events <- extract_environment_array(
  df_rounds_raw,
  "environment_context_media_events"
)

df_external_actions <- extract_environment_array(
  df_rounds_raw,
  "environment_context_external_actor_actions"
)

df_social_alerts <- extract_environment_array(
  df_rounds_raw,
  "environment_context_social_manager_alerts"
)

df_agents_unavailable <- extract_environment_array(
  df_rounds_raw,
  "environment_context_agents_unavailable"
)

df_critical_deadlines <- extract_environment_array(
  df_rounds_raw,
  "environment_context_critical_deadlines"
)

df_news <- extract_environment_array(
  df_rounds_raw,
  "environment_context_news"
)
```

## Dataset inventory

The following table confirms the main analytical tables created from the raw dataset.

```{r}
dataset_inventory <- list(
  "Participants" = df_participants,
  "Communications" = communications,
  "Environment and market context" = df_environment_clean,
  "Media events" = df_media_events,
  "External actor actions" = df_external_actions,
  "Social manager alerts" = df_social_alerts,
  "Unavailable agents" = df_agents_unavailable,
  "Critical deadlines" = df_critical_deadlines,
  "News" = df_news
) %>%
  purrr::imap_dfr(~ tibble(
    Dataset = .y,
    Rows = nrow(.x),
    Columns = ncol(.x)
  ))

knitr::kable(dataset_inventory)
```

# Data quality checks

## Communication data completeness

The communication table is the main source for the subsequent temporal, behavioural, text and network analysis. Missing values are summarised before features are engineered.

```{r}
missing_summary <- communications %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Count"
  ) %>%
  arrange(desc(Missing_Count))

knitr::kable(missing_summary)
```

## Message identifier and timestamp checks

Message IDs are checked for potential duplicates. Timestamps are also checked after conversion so that any non-parsed values can be reviewed before downstream analysis.

```{r}
duplicate_message_ids <- communications %>%
  filter(!is.na(message_id)) %>%
  count(message_id, sort = TRUE) %>%
  filter(n > 1)

communication_timestamp_check <- communications %>%
  transmute(timestamp_parsed = lubridate::ymd_hms(timestamp, quiet = TRUE)) %>%
  summarise(
    Total_Records = n(),
    Missing_or_Unparsed_Timestamps = sum(is.na(timestamp_parsed))
  )

duplicate_message_id_check <- tibble(
  Check = "Duplicate message IDs",
  Value = nrow(duplicate_message_ids)
)

knitr::kable(communication_timestamp_check)
knitr::kable(duplicate_message_id_check)
```

# Communication feature engineering

The following features prepare the message-level data for later investigation of event sequence, communication exposure, behavioural signals and possible dissemination pathways. These are screening features only; they do not by themselves establish intent or wrongdoing.

## Temporal features

The inappropriate release occurred on 5 June 2046. The features below distinguish normal activity from the escalation and critical windows immediately before the breach.

```{r}
breach_date <- lubridate::ymd("2046-06-05")

communications_clean <- communications %>%
  mutate(
    timestamp = lubridate::ymd_hms(timestamp, quiet = TRUE),
    round_hour = lubridate::ymd_hms(round_hour, quiet = TRUE)
  ) %>%
  arrange(timestamp) %>%
  mutate(
    communication_date = as_date(timestamp),
    communication_hour = hour(timestamp),
    communication_day = wday(timestamp, label = TRUE),

    communication_period = case_when(
      communication_hour >= 6 & communication_hour < 12 ~ "Morning",
      communication_hour >= 12 & communication_hour < 18 ~ "Afternoon",
      communication_hour >= 18 & communication_hour < 24 ~ "Evening",
      TRUE ~ "Late Night"
    ),

    days_before_embargo_breach = as.integer(breach_date - communication_date),

    near_breach_window = case_when(
      days_before_embargo_breach < 0 ~ "Post-breach",
      days_before_embargo_breach <= 2 ~ "Critical Period",
      days_before_embargo_breach <= 5 ~ "Escalation Period",
      TRUE ~ "Normal Period"
    ),

    message_sequence = row_number()
  )

knitr::kable(
  communications_clean %>%
    select(
      timestamp, communication_date, communication_hour,
      communication_period, days_before_embargo_breach,
      near_breach_window, message_sequence
    ) %>%
    head(10)
)
```

## Communication scope and exposure risk

Channels are grouped into broader communication scopes. This allows later analyses to distinguish public communication from private, internal and side-channel exchanges.

```{r}
communications_clean <- communications_clean %>%
  mutate(
    communication_scope = case_when(
      channel %in% c("official_post", "personal_post") ~ "Public",
      channel %in% c("one_on_one_chat") ~ "Private",
      channel %in% c("side_huddle") ~ "Side-channel",
      channel %in% c("comms_huddle") ~ "Internal Group",
      TRUE ~ "Other"
    ),

    exposure_risk_level = case_when(
      communication_scope == "Public" ~ "High Exposure",
      communication_scope == "Side-channel" ~ "Moderate Exposure",
      communication_scope == "Private" ~ "Low Exposure",
      communication_scope == "Internal Group" ~ "Controlled Exposure",
      TRUE ~ "Unknown"
    ),

    is_public_channel = communication_scope == "Public"
  )

channel_summary <- communications_clean %>%
  count(channel, communication_scope, exposure_risk_level, sort = TRUE)

knitr::kable(channel_summary)
```

## Text and behavioural indicators

Message content and any available internal-state text are combined into one analysis field. Keyword indicators then identify potential references to the merger, embargo controls, public disclosure and operational pressure.

```{r}
combine_text_fields <- function(data, columns) {
  if (length(columns) == 0) {
    return(rep("", nrow(data)))
  }

  apply(data[columns], 1, function(x) {
    stringr::str_squish(
      stringr::str_c(replace_na(as.character(x), ""), collapse = " ")
    )
  })
}

internal_state_columns <- names(communications_clean) %>%
  stringr::str_subset("^internal_state_")

communications_clean$internal_state_text <- combine_text_fields(
  communications_clean,
  internal_state_columns
)

communications_clean <- communications_clean %>%
  mutate(
    content_lower = str_to_lower(replace_na(content, "")),
    analysis_text = str_squish(
      str_c(content_lower, str_to_lower(internal_state_text), sep = " ")
    ),

    contains_project_terms = str_detect(
      analysis_text,
      "harborcrest|civicloom|merger|rebrand|capital|deal"
    ),

    contains_embargo_terms = str_detect(
      analysis_text,
      "embargo|judge|compliance|approved|blocked|deadline"
    ),

    contains_public_exposure_terms = str_detect(
      analysis_text,
      "public|post|fle?x|release|leak|disclose|announcement"
    ),

    contains_pressure_terms = str_detect(
      analysis_text,
      "pressure|urgent|concern|risk|legal|scrutiny|journal|algorithmiceviction"
    ),

    has_internal_reasoning = str_length(internal_state_text) > 0,
    message_length = str_length(content_lower),
    word_count = str_count(content_lower, "\\S+"),

    risk_signal_score =
      as.integer(contains_project_terms) +
      as.integer(contains_embargo_terms) +
      as.integer(contains_public_exposure_terms) +
      as.integer(contains_pressure_terms),

    risk_signal_group = case_when(
      risk_signal_score >= 3 ~ "High Signal",
      risk_signal_score == 2 ~ "Moderate Signal",
      risk_signal_score == 1 ~ "Low Signal",
      TRUE ~ "No Signal"
    )
  )

risk_signal_summary <- communications_clean %>%
  count(near_breach_window, risk_signal_group, sort = TRUE)

knitr::kable(risk_signal_summary)
```

## Agent-level preparation table

The message-level features are summarised by agent to provide a reusable table for later comparison of activity, exposure and risk signals.

```{r}
agent_feature_summary <- communications_clean %>%
  group_by(agent_id, agent_role) %>%
  summarise(
    total_messages = n(),
    public_messages = sum(is_public_channel, na.rm = TRUE),
    high_signal_messages = sum(risk_signal_group == "High Signal", na.rm = TRUE),
    average_risk_signal_score = mean(risk_signal_score, na.rm = TRUE),
    messages_with_internal_reasoning = sum(has_internal_reasoning, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(high_signal_messages), desc(total_messages))

knitr::kable(head(agent_feature_summary, 15))
```

# Preparing network-ready communication tables

## Directed communication edges

The recipient field may contain several recipients. It is expanded so that each sender-recipient pair becomes one directed communication edge. Broadcast messages to `ALL` are excluded because they do not identify an individual receiver.

```{r}
communication_edges <- communications_clean %>%
  separate_rows(recipients, sep = "\\s*,\\s*") %>%
  transmute(
    from = str_trim(agent_id),
    to = str_trim(recipients),
    channel,
    communication_scope,
    exposure_risk_level,
    risk_signal_group,
    timestamp
  ) %>%
  filter(
    !is.na(from), !is.na(to),
    from != "", to != "", to != "ALL",
    from != to
  )

show_table(
  communication_edges %>%
    select(from, to, channel, communication_scope, exposure_risk_level,
           risk_signal_group, timestamp),
  text_columns = c("from", "to"),
  max_chars = 45
)
```

## Nodes and weighted edges

Nodes represent all unique senders and recipients. Weighted edges aggregate repeated sender-recipient interactions and retain communication scope and risk classification.

```{r}
communication_nodes <- tibble(
  name = sort(unique(c(communication_edges$from, communication_edges$to)))
) %>%
  left_join(
    communications_clean %>%
      distinct(agent_id, agent_role) %>%
      rename(name = agent_id, role = agent_role),
    by = "name"
  ) %>%
  mutate(role = replace_na(role, "Recipient / System"))

communication_edges_weighted <- communication_edges %>%
  count(
    from,
    to,
    communication_scope,
    exposure_risk_level,
    risk_signal_group,
    name = "weight"
  ) %>%
  mutate(
    interaction_intensity = case_when(
      weight >= 10 ~ "High Interaction",
      weight >= 5 ~ "Moderate Interaction",
      TRUE ~ "Low Interaction"
    )
  )

network_inventory <- tibble(
  Measure = c(
    "Communication nodes",
    "Directed communication edges",
    "Weighted communication edges"
  ),
  Value = c(
    nrow(communication_nodes),
    nrow(communication_edges),
    nrow(communication_edges_weighted)
  )
)

knitr::kable(network_inventory)
show_table(communication_edges_weighted, max_chars = 45)
```

## Creating the communication network object

The node and edge tables are combined into a directed network object. It is saved for later analysis or visualisation but is not plotted on this data-preparation page.

```{r}
communication_network <- tidygraph::tbl_graph(
  nodes = communication_nodes,
  edges = communication_edges_weighted,
  directed = TRUE
)

network_object_summary <- tibble(
  Measure = c("Network type", "Nodes", "Weighted edges"),
  Value = c("Directed communication network", nrow(communication_nodes),
            nrow(communication_edges_weighted))
)

knitr::kable(network_object_summary)
```

# Saving prepared data

All cleaned tables and engineered features are saved as one RDS object. The other Quarto pages and the Shiny application can load this prepared file directly rather than repeatedly unpacking the nested source RDS object.

```{r}
prepared_mc1 <- list(
  participants = df_participants,
  communications = communications_clean,
  environment = df_environment_clean,
  media_events = df_media_events,
  external_actions = df_external_actions,
  social_alerts = df_social_alerts,
  agents_unavailable = df_agents_unavailable,
  critical_deadlines = df_critical_deadlines,
  news = df_news,
  agent_feature_summary = agent_feature_summary,
  communication_edges = communication_edges,
  communication_nodes = communication_nodes,
  communication_edges_weighted = communication_edges_weighted,
  communication_network = communication_network
)

output_path <- file.path("data", "MC1", "mc1_prepared_data.rds")
dir.create(dirname(output_path), recursive = TRUE, showWarnings = FALSE)
readr::write_rds(prepared_mc1, output_path)

knitr::kable(tibble(
  Output = output_path,
  Saved_Objects = length(prepared_mc1)
))
```

A later page can load the prepared object using:

```r
prepared_mc1 <- readr::read_rds(
  file.path("data", "MC1", "mc1_prepared_data.rds")
)
```
